Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-25 paper/paper.tex
Take a pass and fix minor things

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type theories.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.
\scw{need one sentence to motivate why logical relations are important}

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that proofs that use
logical relations are not applicable.  \scw{Well, actually, we did use a
  step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys'
  logical fragment is simple type + identity so I don't think that counts..}
But this is not the case. Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code, these definitions are easy to miss and difficult to
isolate from their surrounding contexts. As a result, they are not accessible
to casual readers.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. \scw{Say that we don't have impredicativity, but neither does anyone
  else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
 Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of $[[G]]$.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Par} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Par} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

@@

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the existence of a
fixed point with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the inversion
principle (Lemma~\ref{lemma:piinvalt}) we derive after proving functionality.

@@

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

\section{Existence of $\beta$-normal forms}
\label{sec:extension}
In this section, we show how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of
$\beta$ normal forms for (open and closed) well-typed terms.
\scw{Why is this a sketch? We've done the whole proof and we will
explain it all, right?}\yl{s/sketch out/show. I used sketch because we
didn't spell out the full proofs}

@@

\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}

@@

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}

\begin{lemma}[Wn extensionality]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

  The \rref{I-Set} case is the most involved. We need to show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  $[[j]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn Pi A B]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that $[[var 0]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn B {var 0}]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

@@

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.
The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.

@@

Due to the non-deterministic nature of parallel reduction, we need to take a
few more steps to convert the existence of $\beta$-normal form into a decision
procedure for type conversion. More specifically, we can show that a
deterministic evaluation strategy such as leftmost-outermost reduction can
always find the $\beta$-normal form if there exists one. Then the termination
of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
However, we omit such proofs since they can can be formulated on untyped
lambda terms and thus are orthogonal to the specifics of dependently typed
systems. Instead, we redirect readers to \citet{factorization-essentially,
  takahashi-parallel-reduction} for the details.

\scw{Name and foreshadow these lemmas when they are introduced so that
it is easier to refer to them in the discussion.}
Finally, we want to point out that
the adequacy property (Lemma~\ref{lemma:adequacy}) are in fact required in the normalization proof
for simply typed languages~\citep{abel2019poplmark}. Similar to the
simply typed scenario, adequacy needs to be proven before the
fundamental theorem so we can handle elimination rules such as
\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. % In
% \citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
% in the exact same way to show the normalization of lambda forms
Dependent types make the adequacy proof slightly more complicated
as we also need to know that every type has a normal form, not just
the terms.
This complicates our proof specifically in the \rref{I-Set} case for
our adequacy property (Lemma~\ref{lemma:adequacy}).

>>>>

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type
theories.
Tutorial material on this topic~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types, pierce2004advanced,harper2016practical} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.
\scw{need one sentence to motivate why logical relations are important}

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that proofs that use
logical relations are not applicable.
But this is not the case. Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code, these definitions are easy to miss and difficult to
isolate from their surrounding contexts. As a result, they are not accessible
to casual readers.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. % \scw{Say that we don't have impredicativity, but neither does anyone
  % else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
 Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of
$[[G]]$. In \rref{T-J}, the elimination form for the idnetity type,
$[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
to refer to the same binders.

@@

% To derive consistency, it suffices to restrict $[[S]]$ to the set of
% terms that reduce to closed terms.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type corresponds to terms that
evaluate to $[[true]]$ or $[[false]]$. However, the characterization
of $[[Bool]]$ in our logical relation does not always correspond to
well-typed or even closed terms. For example, we have $[[if false var 0
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evalutes to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Par} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Par} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

@@

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.

@@

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
\begin{lemma}[Pi Inversion Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

\section{Existence of $\beta$-normal forms}
\label{sec:extension}
In this section, we show how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of
$\beta$ normal forms for (open and closed) well-typed terms.

@@

\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation (new and changed rules)}
  \label{fig:logrelopen}
\end{figure}

@@

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes,
parallel reduction can not take any real reduction steps and can only
step into a term itself. However, for the purpose of our proof,
Lemma~\ref{lemma:parnenf} is sufficient. Furthermore,
Lemma~\ref{lemma:parnenf}, in its current weaker form, would still hold after we extend our vequational
theory with the function $\eta$ rule, where parallel reduction can
take $\eta$ steps but still preserves $\beta\eta$-normal form.

\begin{lemma}[Wn extensionality]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemmas~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

  The \rref{I-Set} case is the most involved. We need to show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn Pi A B]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that $[[var 0]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn B {var 0}]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

@@

The formulation of the valuation and semantic well-typedness from
Section~\ref{sec:logrelproof} and the fundamental lemma remains
unchanged.
The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.

@@

\yl{commented out the following paragraph because I accidentally wrote
the same thing twice. Will do a comparison and if this is better
written I'll replace the part from the $\beta\eta$ section with it}
% Due to the non-deterministic nature of parallel reduction, we need to take a
% few more steps to convert the existence of $\beta$-normal form into a decision
% procedure for type conversion. More specifically, we can show that a
% deterministic evaluation strategy such as leftmost-outermost reduction can
% always find the $\beta$-normal form if there exists one. Then the termination
% of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
% However, we omit such proofs since they can can be formulated on untyped
% lambda terms and thus are orthogonal to the specifics of dependently typed
% systems. Instead, we redirect readers to \citet{factorization-essentially,
%   takahashi-parallel-reduction} for the details.

Finally, we want to point out that
the adequacy property (Lemma~\ref{lemma:adequacy}) are in fact required in the normalization proof
for simply typed languages~\citep{abel2019poplmark}. Similar to the
simply typed scenario, adequacy needs to be proven before the
fundamental theorem so we can handle elimination rules such as
\rref{T-If} and \rref{T-App} where the scrutinee is a neutral term. % In
% \citet{abel2019poplmark}, a variant of Lemma~\ref{lemma:extwn} is used
% in the exact same way to show the normalization of lambda forms
Dependent types make the adequacy proof slightly more complicated
as we also need to know that every type has a normal form, not just
the terms.
This complicates our proof specifically in the \rref{I-Set} case for
our adequacy property (Lemma~\ref{lemma:adequacy}).
